\chapter{Conclusions and Outlook}
\label{chap:conclusion}

\section{Summary}

This handbook documents two tracks for piecewise PV inversion of the
2025-01-08 California blocking event:

\begin{enumerate}\setlength\itemsep{0pt}
  \item the \textbf{Wu Fortran reference} (Part~I,
        \Cref{chap:wu_solver,chap:wu_pipeline}), a finite-difference SOR
        solver that runs end-to-end and reproduces the Davis 2022
        Fig.~8 diagnostic.  It is the project \emph{benchmark};
  \item the \textbf{Python spherical-harmonic BALNC port} (Part~II,
        \Cref{chap:wu_balnc}), a faithful re-implementation of Wu's
        non-dimensional balance system on the hemispheric SH grid,
        built on \texttt{pvtend.sh\_ops}.
\end{enumerate}

\noindent Both tracks share the mathematical foundations of
\Cref{chap:math} (Ertel PV, hydrostatic and balance closures,
linearization) and the SH operator and forward-PV machinery of
\Cref{chap:sh,chap:forward}.

\section{Status and Known Limitations}

\begin{itemize}\setlength\itemsep{0pt}
  \item \textbf{Wu reference --- working.}  Pass~A/B/C/D complete and
        the Fig.~8 replica (\Cref{fig:wu_fig8}) reproduces the
        published structure.  For this case the total-balance Pass~C is
        a consistency check: it leaves the \texttt{pvpialln} field
        unchanged to solver tolerance.
  \item \textbf{SH port --- BALNC does not yet converge.}  With the
        explicit nonlinear terms frozen, the streamfunction norm drifts
        upward under a slow $H$--$\psi$ linear feedback rather than
        settling (\Cref{sec:wu_open}).  The piecewise BALP pass is
        therefore not yet run.
  \item \textbf{$f_0$-plane closure.}  The legacy linear closure
        $\dPhi\approx f_0\dpsi$ (\Cref{sec:closure}), still cited by the
        SH port, under-estimates upper-level geopotential and accounts
        for an $\approx\!0.1\times$ amplitude bias relative to the Wu
        reference.
\end{itemize}

\section{Next Steps}

\begin{enumerate}\setlength\itemsep{0pt}
  \item \textbf{Converge BALNC (Pass~C) in the SH port.}  Candidate
        fixes (\Cref{sec:wu_open}): solve the $H$--$\psi$ system
        \emph{coupled} rather than alternating; treat the vertical
        $BB/BH/BL$ coupling \emph{implicitly} (tridiagonal in $k$);
        relax the first $H$ update.
  \item \textbf{Run the SH BALP pass} (\texttt{balp.py}) once Pass~C is
        stable, inlining the same $n{\le}1$ mode projection.
  \item \textbf{Quantitative validation} of the SH port against the Wu
        Fortran benchmark (\Cref{fig:wu_fig8}), once it converges.
\end{enumerate}

