\chapter{The Wu Fortran Reference Solver}
\label{chap:wu_solver}

This chapter describes the working benchmark: W.~Wu's finite-difference
piecewise PV-inversion code, ported only in the sense of being
\emph{driven} from Python (\texttt{wu/steps/}) while the numerical core
remains the original Fortran-77.  The three programs live in
\texttt{wu/fortran/} and are compiled with
\texttt{gfortran -std=legacy -O2 -fno-automatic}.

\section{Grid and Vertical Levels}
\label{sec:wu_grid}

The solver operates on a bounded Northern-Hemisphere window
(\texttt{config.py}):

\begin{center}\small
\begin{tabular}{lll}
  \toprule
  Quantity & Symbol / code & Value \\
  \midrule
  Longitudes & $N_X$ & $87$ \\
  Latitudes  & $N_Y$ & $51$ \\
  Levels     & $N_W$ & $10$ \\
  Latitude row & $\phi_i = 85.5 - 1.5\,i$ & $85.5^\circ\!\to\!-13.5^\circ$ (N$\to$S) \\
  Longitude col & $\lambda_j = -169.5 + 1.5\,j$ & $-169.5^\circ\!\to\!-40.5^\circ$ \\
  Pressure levels & PLEVS [hPa] & $1000,925,850,700,600,$ \\
                  &             & $500,400,300,250,200$ \\
  Earth radius & $R_{\oplus}=2\times10^{7}/\pi$ & $6.366\times10^{6}$ m \\
  Grid spacing & $\Delta p=\Delta l = R_{\oplus}\,\mathrm{rad}(1.5^\circ)$ & $1.667\times10^{5}$ m \\
  \bottomrule
\end{tabular}
\end{center}

\begin{warnbox}[Fortran \texttt{PARAMETER} blocks are hard-wired]
$N_X,N_Y$ appear as compile-time \texttt{PARAMETER}s inside every
\texttt{.f} file.  Changing the grid requires editing and recompiling
the Fortran, not just \texttt{config.py}.  The Python drivers pass
solver \emph{parameters} (relaxation factors, iteration caps) over
\texttt{stdin}; they cannot change the array dimensions.
\end{warnbox}

\section{The Three Programs and Four Passes}
\label{sec:wu_passes}

\begin{center}\small
\begin{tabular}{llll}
  \toprule
  Pass & Program & Subroutine & Produces \\
  \midrule
  A & \texttt{pvpialln}  & ---    & \texttt{meanq}, \texttt{meanh} (mean PV, $\psi$, $\Phi$) \\
  B & \texttt{pvpialln}  & ---    & \texttt{event\_q.out}, \texttt{event\_h.out} (event PV, $\psi$, $\Phi$) \\
  C & \texttt{qinvert21} & BALNC  & \texttt{event\_bal.out} (total balanced $\psi$, $\Phi$) \\
  D & \texttt{qinvertp21}& BALP   & \texttt{event\_pert.out} (3-piece perturbation $\psi'$) \\
  \bottomrule
\end{tabular}
\end{center}

\paragraph{Pass A/B --- \texttt{pvpialln}.}  Computes Ertel PV on the
interior $\sigma$-levels ($K=2\ldots N_W{-}1$, i.e.\ $925$--$250$ hPa),
the balanced streamfunction $\psi$ by relative-vorticity inversion, the
geopotential $\Phi$ on pseudo-height, and the boundary potential
temperature $\theta$ at top and bottom.  Pass A is run on the
11-day climatological mean state; Pass B on the event state.

\paragraph{Pass C --- \texttt{qinvert21} (BALNC).}  Solves the full
\emph{total} balanced inversion: a 2-D Helmholtz equation for $\psi$ on
each $\sigma$-level coupled to a 3-D PV--$\psi$ relation under
gradient-wind balance, by successive over-relaxation with
under-relaxation between $\psi$ and $\Phi$.  The non-dimensional form of
this coupled system --- which the Python SH port re-implements verbatim
--- is given in \Cref{sec:wu_balnc_eqs}.

\paragraph{Pass D --- \texttt{qinvertp21} (BALP).}  Partitions the PV
\emph{perturbation} $\dq=q_{\text{event}}-q_{\text{mean}}$ into three
vertical pieces and inverts each independently to its perturbation
streamfunction $\psi'$:

\begin{center}\small
\begin{tabular}{lll}
  \toprule
  Piece & Levels $K$ & Pressures \\
  \midrule
  1 (lower)  & $\{1,2\}$       & $1000, 925$ hPa \\
  2 (middle) & $\{3,4\}$       & $850, 700$ hPa \\
  3 (upper)  & $\{5,6,7,8,9\}$ & $600,500,400,300,250$ hPa \\
  \bottomrule
\end{tabular}
\end{center}

\section{Solver Parameters}
\label{sec:wu_params}

The relaxation parameters below are passed over \texttt{stdin} by the
step drivers and are the values that make Pass~C/D stable on this grid:

\begin{center}\small
\begin{tabular}{lll}
  \toprule
  Parameter & Value & Role \\
  \midrule
  \texttt{OMEGS} & $1.4$ & SOR factor for the $\psi$ (streamfunction) sweep \\
  \texttt{OMEGH} & $1.4$ & SOR factor for the $\Phi$/$H$ (geopotential) sweep \\
  \texttt{PART}  & $0.5$ & under-relaxation coupling $\psi\leftrightarrow\Phi$ \\
  \texttt{INLIN} & $0$   & \textbf{linear} balance (mandatory; see warning) \\
  \texttt{MAXT}  & $200$ & outer-iteration cap \\
  \texttt{THRSH} & $0.01$& convergence threshold \\
  \bottomrule
\end{tabular}
\end{center}

\begin{warnbox}[\texttt{INLIN=0} is mandatory]
With the nonlinear balance enabled (\texttt{INLIN=1}) the iteration
explodes at the $250$\,hPa jet.  All production passes use the
\emph{linear} balance \texttt{INLIN=0}.  The same instability re-appears
as the dominant obstacle in the SH port (\Cref{sec:wu_open}).
\end{warnbox}

\begin{keybox}[Wu's PV units are solver-internal]
Wu's internal $Q$ is \emph{not} in PVU; it is roughly $600\times$ a PVU
(an artifact of the non-dimensionalisation, \Cref{sec:wu_nondim}).
Every physical-unit diagnostic (PV anomaly contours, PV advection) is
therefore computed from a \emph{separately} evaluated Ertel PV on the
ERA5 fields (\Cref{sec:step08}), never from Wu's $Q$ directly.
\end{keybox}
